---
title: "Creating Fake Data for Plot the Model using the GRCBDS simulator"
author: "Jeffrey Walker"
date: "`r Sys.Date()`"
output: html_document
---

# Background

The fake data generated to demonstrate Plot the Model is motivated by

[Xu, Zhisheng, et al. "Disuse-associated loss of the protease LONP1 in muscle impairs mitochondrial function and causes reduced skeletal muscle mass and strength." Nature Communications 13.1 (2022): 894.](https://www.nature.com/articles/s41467-022-28557-5)

LONP1 is a protease (a protein that breaks down other proteins) expressed in the mitochondrial matrix (the fluid of the inner compartment of the mitochondria). Muscle atrophy due to disuse is correlated with lower LONP1 expression. Does LONP1 maintain muscle health (size and strength)?

```{r setup, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE)

# wrangling packages
library(here) # here makes a project transportable
library(janitor) # clean_names
library(readxl) # read excel, duh!
library(writexl) # read excel, duh!
library(data.table) # magical data frames
library(magrittr) # pipes
library(stringr) # string functions
library(forcats) # factor functions

# analysis packages
library(emmeans) # the workhorse for inference
library(nlme) # gls and some lmm
library(lme4) # linear mixed models
library(lmerTest) # linear mixed model inference
library(afex) # ANOVA linear models
library(glmmTMB) # generalized linear models
library(MASS) # negative binomial and some other functions
library(car) # model checking and ANOVA
library(DHARMa) # model checking
library(mvtnorm)
library(MHTdiscrete) # sidak

# graphing packages
library(ggsci) # color palettes
library(ggpubr) # publication quality plots
library(ggforce) # better jitter
library(cowplot) # combine plots
library(knitr) # kable tables
library(kableExtra) # kable_styling tables

# ggplot_the_model.R packages not loaded above
library(insight)
library(lazyWeave)

# use here from the here package
here <- here::here
# use clean_names from the janitor package
clean_names <- janitor::clean_names
# use transpose from data.table
transpose <- data.table::transpose

# load functions used by this text written by me
# ggplot_the_model.R needs to be in the folder "R"
# if you didn't download this and add to your R folder in your
# project, then this line will cause an error
source_path <- here("R", "ggplot_the_model.R")
source(source_path)

data_folder <- "data"
image_folder <- "images"
output_folder <- "output"
```

```{r}
glmm_simulator <- function(
  seed_i = 1,
  family = "gaussian",
  n_sim = 1, # number of simulations
  n_treat = 3, # tau
  n_block = 6, # beta_1, number of (nested) blocks
  n_rep = 1, # eta, number of experimental replicates
  n_ss = 1, # epsilon, number of subsamples within each block:treatment.
  n_exp = 1, # number of experiments
  design = "blocked"
    # if "crd", then completely randomized design
    # if "blocked", then all treatments within block.
    # if "pseudoreplicated", then single treatment per block and all replicates within block are subsamples.
){
  
  # build model matrix
  
}
```

# CRD - 2 x 1 (t-test)

The researchers want to estimate the effect of LONP1 on muscle cross-sectional area (CSA). For the experiment, the researchers randomly sample 12 female mice, each from a different litter, and randomly assign 6 mice to Control (with functional LONP1) and 6 mice to mKO (muscle specific LONP1 knock-out).

This is a Completely Randomized Design (CRD) with a single factor (Treatment) with two levels (Control and mKO). The classic test for this is Student's (or Independent Samples) t-Test, which is a special case of a Linear Model.

For PlotTheModel, use

1. Experimental Design = CRD
2. 

```{r fake-2x1-crd}
seed <- 1
fake_data <- simulator(
  seed_i = seed,
  n_sim = 1,
  n_treat = 2, # number of treatment levels
  n_block = 1, # number of litters (blocks)
  n_rep = 6, # number of mice per litter:treatment
  n_ss = 1, # number of subsamples
  n_exp = 1, # number of experiments
  design = "pseudoreplicated",
  correlated_slopes = FALSE, # TRUE uses random int/slope model to generate data
  beta = c(2400, -800),
  sigma_exp.block = 0, # only sampling 1 litter
  sigma_rep = 400, # among mouse variance
  sigma_ss = 0, # among technical reps within mouse
  block_name = "litter",
  rep_name = "mouse",
  ss_name = "ss"
)
setnames(fake_data, "sim_1", "csa") # cross-sectional area
fake_data[, genotype := ifelse(treatment == "Cn", "Control", "mKO")]
fake_data[, genotype := factor(genotype, levels = c("Control", "mKO"))]
m1 <- lm(csa ~ genotype, data = fake_data)
m1_emm <- emmeans(m1, specs = "genotype")
m1_pairs <- contrast(m1_emm,
                     method = "revpairwise")
ggplot_the_response(m1, m1_emm, m1_pairs)

outfile_name <- "crd2.xlsx"
outpath <- here(data_folder, outfile_name)
y_cols <- c("mouse", "genotype", "csa")
write_xlsx(fake_data[, .SD, .SDcols = y_cols], outpath)
```

# CRD - 3 x 1 (ANOVA)

```{r fake-3x1-crd}
seed <- 1
fake_data <- simulator(
  seed_i = seed,
  n_sim = 1,
  n_treat = 3, # number of treatment levels
  n_block = 1, # number of litters (blocks)
  n_rep = 6, # number of mice per litter:treatment
  n_ss = 1, # number of subsamples
  n_exp = 1, # number of experiments
  design = "pseudoreplicated",
  correlated_slopes = FALSE, # TRUE uses random int/slope model to generate data
  beta = c(2400, -900, -800),
  sigma_exp.block = 0, # only sampling 1 litter
  sigma_rep = 400, # among mouse variance
  sigma_ss = 0, # among technical reps within mouse.
  block_name = "litter",
  rep_name = "mouse",
  ss_name = "ss"
)
setnames(fake_data, c("treatment", "sim_1"), c("fake_treatment", "csa")) # cross-sectional area
fake_data[fake_treatment == "Cn", treatment := "Control"]
fake_data[fake_treatment == "Tr1", treatment := "Denerv"]
fake_data[fake_treatment == "Tr2", treatment := "mKO"]
fake_data[, treatment := factor(treatment, levels = c("Control", "Denerv", "mKO"))]
m1 <- lm(csa ~ treatment, data = fake_data)
m1_emm <- emmeans(m1, specs = "treatment")
m1_pairs <- contrast(m1_emm,
                     method = "revpairwise") |>
  summary(infer = TRUE)
ggplot_the_response(m1, m1_emm, m1_pairs)

outfile_name <- "crd3.xlsx"
outpath <- here(data_folder, outfile_name)
y_cols <- c("mouse", "treatment", "csa")
write_xlsx(fake_data[, .SD, .SDcols = y_cols], outpath)
```


# CRD - 2 x 2 (Twoway ANOVA)

```{r fake-2x2-crd}
seed <- 1
fake_data <- simulator(
  seed_i = seed,
  n_sim = 1,
  n_treat = 4, # number of treatment levels
  n_block = 1, # number of litters (blocks)
  n_rep = 6, # number of mice per litter:treatment
  n_ss = 1, # number of subsamples
  n_exp = 1, # number of experiments
  design = "pseudoreplicated",
  correlated_slopes = FALSE, # TRUE uses random int/slope model to generate data
  beta = c(2400, -800, -700, -1200),
  sigma_exp.block = 0, # only sampling 1 litter
  sigma_rep = 400, # among mouse variance
  sigma_ss = 0, # among technical reps within mouse
  block_name = "litter",
  rep_name = "mouse",
  ss_name = "ss"
)
setnames(fake_data, c("treatment", "sim_1"), c("fake_treatment", "csa")) # cross-sectional area
fake_data[fake_treatment == "Cn", treatment := "Intact\nLONP1"]
fake_data[fake_treatment == "Tr1", treatment := "Denerv\nLONP1"]
fake_data[fake_treatment == "Tr2", treatment := "Intact\nmKO"]
fake_data[fake_treatment == "Tr3", treatment := "Denerv\nmKO"]
fake_data[, c("surgery", "genotype") := tstrsplit(treatment, "\n")]
fake_data[, treatment := factor(treatment, levels = c("Intact\nLONP1", "Denerv\nLONP1", "Intact\nmKO", "Denerv\nmKO"))]
m1 <- lm(csa ~ treatment, data = fake_data)
m1_emm <- emmeans(m1, specs = "treatment")
m1_pairs <- contrast(m1_emm,
                     method = "revpairwise")
ggplot_the_response(m1, m1_emm, m1_pairs, dots = "sina", jitter_width = 0.5)
ggplot_the_response(m1, m1_emm, m1_pairs, dots = "jitter", jitter_width = 0.2)

fake_data[, surgery := factor(surgery, levels = c("Intact", "Denerv"))]
fake_data[, genotype := factor(genotype, levels = c("LONP1", "mKO"))]
m1 <- lm(csa ~ surgery * genotype, data = fake_data)
m1_emm <- emmeans(m1, specs = c("surgery", "genotype"))
m1_pairs <- contrast(m1_emm,
                     method = "revpairwise") |>
  summary(infer = TRUE)


outfile_name <- "crd2x2.xlsx"
outpath <- here(data_folder, outfile_name)
y_cols <- c("mouse", "surgery", "genotype", "treatment", "csa")
write_xlsx(fake_data[, .SD, .SDcols = y_cols], outpath)
```

# CRDS - 2 x 1 (Nested)

Here the response is myotube diameter. 80 myotubes per mouse were measured. See fig5h

```{r fake-crds}
seed <- 1
# sigma_e^2 = sigma_rep^2 + sigma_ss^2
# icc = sigma_rep^2 / sigma_e^2
icc <- 0.6
sigma_e <- 2
sigma_rep <- sqrt(icc*sigma_e^2)
sigma_ss <- sqrt(sigma_e^2 - sigma_rep^2)
fake_data <- simulator(
  seed_i = seed,
  n_sim = 1,
  n_treat = 2, # number of treatment levels
  n_block = 1, # number of litters (blocks)
  n_rep = 6, # number of mice per litter:treatment
  n_ss = 80, # number of subsamples
  n_exp = 1, # number of experiments
  design = "pseudoreplicated",
  correlated_slopes = FALSE, # TRUE uses random int/slope model to generate data
  beta = c(21, -3),
  sigma_exp.block = 0, # only sampling 1 litter
  sigma_rep = sigma_rep, # among mouse variance
  sigma_ss = sigma_ss, # among technical reps within mouse
  block_name = "litter",
  rep_name = "mouse",
  ss_name = "ss"
)
setnames(fake_data, c("ss", "sim_1"), c("technical_rep", "diameter")) # cross-sectional area
fake_data[, genotype := ifelse(treatment == "Cn", "Control", "mKO")]
fake_data[, genotype := factor(genotype, levels = c("Control", "mKO"))]
m1 <- lm(diameter ~ genotype, data = fake_data)
m1_emm <- emmeans(m1, specs = "genotype")
m1_pairs <- contrast(m1_emm,
                     method = "revpairwise")
ggplot_the_response(m1, m1_emm, m1_pairs)

outfile_name <- "crds.xlsx"
outpath <- here(data_folder, outfile_name)
y_cols <- c("mouse", "technical_rep", "genotype", "diameter")
write_xlsx(fake_data[, .SD, .SDcols = y_cols], outpath)
```

# RCBD - 2 x 1 (paired t-test)

```{r rcbd-2x1}
seed <- 1
# sigma_e^2 = sigma_rep^2 + sigma_ss^2
# icc = sigma_rep^2 / sigma_e^2
icc <- 0.6
sigma_e <- 400
sigma_rep <- sqrt(icc*sigma_e^2)
sigma_ss <- sqrt(sigma_e^2 - sigma_rep^2)

fake_data <- simulator(
  seed_i = seed,
  n_sim = 1,
  n_treat = 2, # number of treatment levels
  n_block = 6, # number of litters (blocks)
  n_rep = 1, # number of mice per litter:treatment
  n_ss = 1, # number of subsamples
  n_exp = 1, # number of experiments
  design = "rcbd",
  correlated_slopes = FALSE, # TRUE uses random int/slope model to generate data
  beta = c(2400, -800),
  sigma_exp.block = sigma_rep, # sd among exp:block (or block if n_exp = 1). This is among mouse
  sigma_exp.block.treat = c(0), # sd among exp:block:treat
  sigma_ss = sigma_ss, # sd among subsamples within replication of treatment:block
  block_name = "mouse",
  rep_name = "limb",
  ss_name = "ss"
)
setnames(fake_data, "sim_1", "csa") # cross-sectional area
fake_data[, genotype := ifelse(treatment == "Cn", "Control", "mKO")]
fake_data[, genotype := factor(genotype, levels = c("Control", "mKO"))]
m1 <- lm(csa ~ genotype, data = fake_data)
m1_emm <- emmeans(m1, specs = "genotype")
m1_pairs <- contrast(m1_emm,
                     method = "revpairwise")
t.test(fake_data[genotype == "Control", csa],
       fake_data[genotype == "mKO", csa], paired = TRUE)
ggplot_the_response(m1, m1_emm, m1_pairs)

outfile_name <- "rcbd2x1.xlsx"
outpath <- here(data_folder, outfile_name)
y_cols <- c("mouse", "genotype", "csa")
write_xlsx(fake_data[, .SD, .SDcols = y_cols], outpath)
```

# RCBD - 2 x 2 (RM-ANOVA)

```{r rcbd-2x2}
seed <- 1
# sigma_e^2 = sigma_rep^2 + sigma_ss^2
# icc = sigma_rep^2 / sigma_e^2
icc <- 0.6
sigma_e <- 400
sigma_rep <- sqrt(icc*sigma_e^2)
sigma_ss <- sqrt(sigma_e^2 - sigma_rep^2)

fake_data <- simulator(
  seed_i = seed,
  n_sim = 1,
  n_treat = 4, # number of treatment levels
  n_block = 8, # number of litters (blocks)
  n_rep = 1, # number of mice per litter:treatment
  n_ss = 1, # number of subsamples
  n_exp = 1, # number of experiments
  design = "rcbd",
  correlated_slopes = FALSE, # TRUE uses random int/slope model to generate data
  beta = c(2400, -800, -700, -1200),
  sigma_exp.block = sigma_rep, # sd among exp:block (or block if n_exp = 1). This is among mouse
  sigma_exp.block.treat = 100, # sd among exp:block:treat
  sigma_ss = sigma_ss, # sd among subsamples within replication of treatment:block
  block_name = "litter",
  rep_name = "mouse",
  ss_name = "ss"
)

setnames(fake_data, c("treatment", "sim_1"), c("fake_treatment", "csa")) # cross-sectional area
fake_data[fake_treatment == "Cn", treatment := "Intact LONP1"]
fake_data[fake_treatment == "Tr1", treatment := "Denerv LONP1"]
fake_data[fake_treatment == "Tr2", treatment := "Intact mKO"]
fake_data[fake_treatment == "Tr3", treatment := "Denerv mKO"]
fake_data[, c("surgery", "genotype") := tstrsplit(treatment, " ")]
m1 <- lm(csa ~ surgery * genotype, data = fake_data)
m1_emm <- emmeans(m1, specs = c("surgery", "genotype"))
m1_pairs <- contrast(m1_emm,
                     method = "revpairwise")
ggplot_the_response(m1, m1_emm, m1_pairs)

outfile_name <- "rcbd2x2.xlsx"
outpath <- here(data_folder, outfile_name)
y_cols <- c("litter", "mouse", "surgery", "genotype", "treatment", "csa")
write_xlsx(fake_data[, .SD, .SDcols = y_cols], outpath)
```

# SPD - 2 x 2 (RM-ANOVA)

```{r spd-2x2}
# use settings for rcbd 2 x 1, then change one mouse replicate to either denerv or intact
seed <- 1
# sigma_e^2 = sigma_rep^2 + sigma_ss^2
# icc = sigma_rep^2 / sigma_e^2
icc <- 0.6
sigma_e <- 400
sigma_rep <- sqrt(icc*sigma_e^2)
sigma_ss <- sqrt(sigma_e^2 - sigma_rep^2)

fake_data <- simulator(
  seed_i = seed,
  n_sim = 1,
  n_treat = 2, # number of treatment levels
  n_block = 12, # number of litters (blocks)
  n_rep = 1, # number of mice per litter:treatment
  n_ss = 1, # number of subsamples
  n_exp = 1, # number of experiments
  design = "rcbd",
  correlated_slopes = FALSE, # TRUE uses random int/slope model to generate data
  beta = c(2400, 0), # will add effects below
  sigma_exp.block = sigma_rep, # sd among exp:block (or block if n_exp = 1). This is among mouse
  sigma_exp.block.treat = c(0), # sd among exp:block:treat
  sigma_ss = sigma_ss, # sd among subsamples within replication of treatment:block
  block_name = "mouse",
  rep_name = "limb",
  ss_name = "ss"
)
setnames(fake_data, "sim_1", "csa") # cross-sectional area
fake_data[, genotype := ifelse(treatment == "Cn", "Control", "mKO")]
fake_data[, genotype := factor(genotype, levels = c("Control", "mKO"))]
# create litter column
fake_data[, litter := str_replace(block, "mouse", "litter")]
# create treatment factor that is a main plot
fake_data[, surgery := rep(rep(c("Intact", "Denerv"), each = 2), 6)]
fake_data[, treatment := paste(surgery, genotype)]
# create treatment (main plot) effect
# beta = c(2400, -800, -700, -1200) these are effects in 2x2 RCBD
fake_data[surgery == "Denerv" & genotype == "Control", csa := csa - 800]
fake_data[surgery == "Intact" & genotype == "mKO", csa := csa - 700]
fake_data[surgery == "Denerv" & genotype == "mKO", csa := csa - 1200]

m1 <- lm(csa ~ surgery * genotype, data = fake_data)
m1_emm <- emmeans(m1, specs = c("surgery", "genotype"))
m1_pairs <- contrast(m1_emm,
                     method = "revpairwise")
ggplot_the_response(m1, m1_emm, m1_pairs)

outfile_name <- "spd2x2.xlsx"
outpath <- here(data_folder, outfile_name)
y_cols <- c("litter", "mouse", "surgery", "genotype", "treatment", "csa")
write_xlsx(fake_data[, .SD, .SDcols = y_cols], outpath)
```

# SPD - 3 x 2 (RM-ANOVA)

# RCBDS - 2 groups

Mouse is the block with Cn in one limb and mKO in the 2nd. As in CRDS, the response is myotube diameter. 80 myotubes per limb per mouse were measured. See fig5h


```{r rcbds-2x1}
seed <- 1
# sigma_e^2 = sigma_rep^2 + sigma_ss^2
# icc = sigma_rep^2 / sigma_e^2
icc <- 0.6
sigma_e <- 2
sigma_rep <- sqrt(icc*sigma_e^2)
sigma_ss <- sqrt(sigma_e^2 - sigma_rep^2)

fake_data <- simulator(
  seed_i = seed,
  n_sim = 1,
  n_treat = 2, # number of treatment levels
  n_block = 6, # number of litters (blocks)
  n_rep = 1, # number of mice per litter:treatment
  n_ss = 2, # number of subsamples
  n_exp = 1, # number of experiments
  design = "rcbd",
  correlated_slopes = FALSE, # TRUE uses random int/slope model to generate data
  beta = c(21, -3),
  sigma_exp.block = sigma_rep, # sd among exp:block (or block if n_exp = 1). This is among mouse
  sigma_exp.block.treat = c(0), # sd among exp:block:treat
  sigma_ss = sigma_ss, # sd among subsamples within replication of treatment:block
  block_name = "mouse",
  rep_name = "limb",
  ss_name = "ss"
)
setnames(fake_data, "sim_1", "diameter") # myotube diameter
fake_data[, genotype := ifelse(treatment == "Cn", "Control", "mKO")]
fake_data[, genotype := factor(genotype, levels = c("Control", "mKO"))]

m1 <- lm(diameter ~ genotype, data = fake_data)
m1_emm <- emmeans(m1, specs = "genotype")
m1_pairs <- contrast(m1_emm,
                     method = "revpairwise") |>
  summary(infer = TRUE)
ggplot_the_response(m1, m1_emm, m1_pairs)

outfile_name <- "rcbds2.xlsx"
outpath <- here(data_folder, outfile_name)
y_cols <- c("mouse", "genotype", "limb", "diameter")
write_xlsx(fake_data[, .SD, .SDcols = y_cols], outpath)
```

# RCBDS - 3 groups

# GRCBD - 2 groups

# GRCBD - 3 groups


# CRD-Gamma - 2 groups

The researchers want to estimate the effect of LONP1 on muscle cross-sectional area (CSA). For the experiment, the researchers randomly sample 12 female mice, each from a different litter, and randomly assign 6 mice to Control (with functional LONP1) and 6 mice to mKO (muscle specific LONP1 knock-out).

This is a Completely Randomized Design (CRD) with a single factor (Treatment) with two levels (Control and mKO). The classic test for this is Student's (or Independent Samples) t-Test, which is a special case of a Linear Model.

For PlotTheModel, use

1. Experimental Design = CRD
2. 


```{r fake-2x0-crd-gamma}

seed <- 1
fake_data <- simulator(
  seed_i = seed,
  n_sim = 1,
  family = "gamma",
  n_treat = 2, # number of treatment levels
  n_block = 1, # number of litters (blocks)
  n_rep = 20, # number of mice per litter:treatment
  n_ss = 1, # number of subsamples
  n_exp = 1, # number of experiments
  design = "pseudoreplicated",
  correlated_slopes = FALSE, # TRUE uses random int/slope model to generate data
  beta = c(2400, -1000),
  sigma_exp.block = 0, # only sampling 1 litter
  sigma_rep = 0, # among mouse variance
  sigma_ss = 1200, # among technical reps within mouse but really this is among mice
  block_name = "litter",
  rep_name = "mouse",
  ss_name = "ss"
)
setnames(fake_data, "sim_1", "csa") # cross-sectional area
fake_data[, genotype := ifelse(treatment == "Cn", "Control", "mKO")]
fake_data[, genotype := factor(genotype, levels = c("Control", "mKO"))]
m1 <- lm(csa ~ genotype, data = fake_data)
m1_emm <- emmeans(m1, specs = "genotype")
m1_pairs <- contrast(m1_emm,
                     method = "revpairwise")
ggplot_the_response(m1, m1_emm, m1_pairs)

glm1 <- glm(csa ~ genotype, family = Gamma(link = "log"), data = fake_data)
glm1_emm <- emmeans(glm1, specs = "genotype", type = "response")
glm1_pairs <- contrast(glm1_emm,
                     method = "revpairwise")

outfile_name <- "crd2x0-gamma.xlsx"
outpath <- here(data_folder, outfile_name)
y_cols <- c("mouse", "genotype", "csa")
write_xlsx(fake_data[, .SD, .SDcols = y_cols], outpath)
```

# CRDS-gamma

```{r}

seed <- 1
# sigma_e^2 = sigma_rep^2 + sigma_ss^2
# icc = sigma_rep^2 / sigma_e^2
icc <- 0.4
sigma_e <- 10
sigma_rep <- sqrt(icc*sigma_e^2)
sigma_ss <- sqrt(sigma_e^2 - sigma_rep^2)
fake_data <- simulator(
  seed_i = seed,
  n_sim = 1,
  family = "gamma",
  n_treat = 2, # number of treatment levels
  n_block = 1, # number of litters (blocks)
  n_rep = 6, # number of mice per litter:treatment
  n_ss = 5, # number of subsamples
  n_exp = 1, # number of experiments
  design = "pseudoreplicated",
  correlated_slopes = FALSE, # TRUE uses random int/slope model to generate data
  beta = c(20, -5),
  sigma_exp.block = 0, # only sampling 1 litter
  sigma_rep = sigma_rep, # among mouse variance
  sigma_ss = sigma_ss, # among technical reps within mouse
  block_name = "litter",
  rep_name = "mouse",
  ss_name = "ss"
)
setnames(fake_data, c("ss", "sim_1"), c("technical_rep", "diameter")) # cross-sectional area
fake_data[, genotype := ifelse(treatment == "Cn", "Control", "mKO")]
fake_data[, genotype := factor(genotype, levels = c("Control", "mKO"))]
m1 <- lm(diameter ~ genotype, data = fake_data)
m1_emm <- emmeans(m1, specs = "genotype")
m1_pairs <- contrast(m1_emm,
                     method = "revpairwise")
ggplot_the_response(m1, m1_emm, m1_pairs)

outfile_name <- "crds-gamma-test.xlsx"
outpath <- here(data_folder, outfile_name)
y_cols <- c("mouse", "technical_rep", "genotype", "diameter")
write_xlsx(fake_data[, .SD, .SDcols = y_cols], outpath)
```

